Ignition of Fuel Beds by Firebrands

### 2022-05-25 conclusions.tex
completed conclusiosns

>>>>

    After the completion of the experimental efforts outlined in the previous chapters the results were aggregated into a single dataset containing the results from the 1086 tests conducted. The parameters included in the model and the corresponding units are as follows:
        \begin{itemize}
            \item Wind speed (\si{\meter\per\second)}
            \item Wind direction relative to longitudinal firebrand axis (\si{\degree})
            \item Ratio of characteristic lengths between particle size and ember size (-)
            \item Chemical composition: mass fractions of Cellulose, Hemicellulose, Lignin, Tannins, and Triglycerides (-)
            \item Fuel bed bulk density (\si{\kilo\gram\per\cubic\meter})
            \item Fuel bed thermal diffusivity (\si{\square\meter\per\second})
            \item Average temperature of the firebrand(s) (\si{\celsius})
            \item Number of firebrands (-)
            \item Spacing between firebrands as multiple of characteristic length (-)
        \end{itemize}
    The aggregated data was then used to create a Random Forest Classifier model using the scikit-learn package in python~\cite{scikit-learn} to estimate ignition outcomes for each test. Using a 0.25/0.75 test train spit on the data the R\textsuperscript{2} validation score for the model with an out of bag score of 0.83. Suggesting that the model accounts can predict the ignition correctly 83\% of the time. While the accuracy of the model is likely acceptable if such a model was to be used as an indicator or predictor of ignition during a fire or to inform preventative measures it would be beneficial to reduce the number of inputs required. An examination of the most important input parameters and refinement of the model found that accuracy was not degraded if number of parameters was reduced to include only the following variables:
        \begin{itemize}
            \item Average temperature of the firebrands
            \item Fuel bed bulk density
            \item Wind speed
            \item Ratio of characteristic lengths between particle size and ember size
            \item Wind direction
        \end{itemize}
    Using the five characteristics above the model, a prediction accuracy of 83\% was maintained. While it is beneficial to have an accurate predictive model for a single set of experiments in a similar apparatus, a model capable of predictions across different is far more valuable for preventing home loss. To evaluate the accuracy of the model across studies the model was applied to a selection of studies from literature. The studies and relevant parameters from each study are shown in Table~\ref{tab:otherStudies}. All of the studies presented in Table~\ref{tab:otherStudies} used hot metal particles dropped onto fine fuel beds of processed material. The studies by Hadden et al. and Urban et al. used pure cellulose as the fuel bed and the study by Zak et al used a blend of grasses processed into a powder. A wind direction of 0\si{\degree} was used for all studies since the geometry of the metal particles is anticipated to produce recirculation zones similar to that of a cylindrical heater parallel to the flow. The average temperature was defined as the average of the initial temperature and ambient temperature. More sophisticated estimates of the average temperature based on the average time to ignition and estimated heat losses from the ember were conducted but the accuracy was not improved. Thus, a simpler metric was chosen.  
        \begin{table}[]
            \centering
            \caption{Data inputs for the model from selected studies.}
            \begin{tabular}{crrrrrc}
                Authors & $\bar{T}$ (\si{\celsius}) &  $\bar{\rho}_{bed}$ (\si{\kilo\gram\per\cubic\meter})& $U$ (\si{\meter\per\second}) & $L_{p}/L_{e}$ (-)& $\Theta_{wind}$ (\si{\degree}) & Citation\\
                \hline
                 Hadden et al. & 260 - 560 & 200 & 0.5 & 0.02 - 0.45 & 0 & \cite{Hadden2011}\\
                 Urban et al. & 285 - 560 & 282 & 0.5 & 0.06 - 0.25 & 0 & \cite{Urban2018}\\
                 Zak et al.    & 310 - 560 & 338 & 0.5 & 0.04 - 0.15 & 0 & \cite{Zak2014}
            \end{tabular}
            \label{tab:otherStudies}
        \end{table}
    The first approach used was to apply reduced model created from only the data generated from studies in this dissertation and predict ignition for each of the studies in literature. The accuracy of the model at predicting ignition for other studies was 50\%, which is entirely unusable for any sort of predictive modelling. However when the model was retrained using the same five parameters but including the Zak et al. data in conjunction with the results of this work the predictive accuracy increased to 96\% for the Hadden et al. study and 81\% for the Urban et al. study. The increase in accuracy suggests that using as few as the five parameters in this model predictions can be made across differences in experimental conditions. Predictions across experimental conditions has been one of the biggest hurdles to creating a general predictive model. The identification of the primary controlling parameters of ignition in this work has created a framework for bridging the predictive gap across various configurations.

### 2022-05-25 manuscript3.tex
completed conclusiosns

\section{Abstract}
\label{sec:abtract3}
    The increasing severity of wildfires and the expansion of the wildland urban interface have placed a greater number of homes at risk for being destroyed by wildfires. Ignition of fuel beds on or near homes by firebrands is a significant source of fire spread and structure loss. Fire prevention and suppression efforts lack an expedient method for estimating the likelihood of ignition for the wide variety of materials around homes.

@@

                    \draw[draw=black, dashed] (3.336, 0) rectangle (4.54, 0.34);
                    \draw[->]         (4.3, 1) -- (4.0, 0.34);
                    \node[right, align=left] at (4.3, 1) {\scriptsize Computational\\ \scriptsize Domain};

@@

            \caption{Diagram of the experimental wind tunnel apparatus. Air flows through the wind tunnel from left to right. The dashed region represents the domain subset used for computational efforts.}

@@

    \section{Conclusions}
    \label{sec:conclusions3}
    Conclusions for this manuscript will go here.
    \cite{MacLean1941}.

>>>>

\section{Abstract}
\label{sec:abtract3}
    The increasing severity of wildfires and the expansion of the wildland urban interface have placed a greater number of homes at risk for being destroyed by wildfires. Ignition of fuel beds on or near homes by firebrands is a significant source of fire spread and structure loss. Fire prevention and suppression efforts lack an expedient method for estimating the likelihood of ignition for the wide variety of materials around homes. In this study the effects of fuel bed chemical composition on the temperature threshold for flaming ignition was considered. Fuel beds of six different species were heated with a cartridge heater as a surrogate firebrand. Tests were conducted in both quiescent and windy conditions. Fuel beds with high lignin content (i.e., Douglas-fir bark and pine park) were not observed to ignite at the maximum temperature at either wind speed. Increasing the wind speed decreased the ignition threshold for Douglas-fir wood and pine wood but increased the threshold for oak wood. Decreases in the ignition threshold are attributed to recirculation zones near the fuel bed and firebrand interface, however the increase in ignition threshold for oak wood suggest that chemical sensitivities may inhibit ignition enhancement due to wind.  This work suggests that areas of high risk of firebrand ignition may reduce ignition risk by using materials high in lignin content.

@@

@@

            \caption{Diagram of the experimental wind tunnel apparatus. Air flows through the wind tunnel from left to right.}

@@

\section{Conclusions}
\label{sec:conclusions3}    
    Flaming ignition tests were conducted for fuel beds of six different materials. Ignition of fuel beds was induced by a cartridge heater that served as a firebrand surrogate. The temperature of the heater was maintained at a fixed temperature throughout each test allowing sensitivities of ignition to fuel bed composition to be evaluated. The ignition probability and chemical composition were used to evaluate the influence of chemical composition of the fuel bed and observe differences in ignition probabilities in the presence of wind and in quiescent conditions. The specific conclusions of this work are as follows:
        \begin{enumerate}
            \item Of the six materials tested only three were able to be ignited within below the 750\si{\celsius} maximum temperature of the firebrand surrogate. Two of the three materials for which ignition was not observed were bark. The higher ignition threshold of the bark materials is attributed to higher lignin concentrations. The resistance to ignition present in these materials suggest that using materials high in lignin in locations at risk for firebrand attack in wildfires may reduce ignitions and subsequent losses of structures.

            \item For Douglas-fir and pine wood, an increase of wind speed decreased the temperature required for ignition. However, for oak wood an increase in wind speed inhibited ignition. Increases in ignition propensity with an increase in wind is attributed to the formation of recirculation zones near the ignition source. In the case of oak wood the formation of recirculation zones inhibits ignition. The root cause of the dichotomy in ignition trends is unclear however differences in the flammability of pyrolysis products corresponding to changes in chemical composition is most likely.

            \item A direct correlation between the fuel bed composition and the ignition propensity was not observed. This is attributed to the complexity of biomass pyrolysis. however, local minimum ignition threshold near the composition of Douglas-fir wood is present that warrants further investigation. 
        \end{enumerate}
    The conclusions of this work show that ignition is sensitive to the chemical composition of fuel beds in both quiescent conditions and under wind. An increase of wind was observed to increase ignition probability for some materials however this was not universally the case. Materials high in lignin were among the least likely to ignite and more suitable for placement in areas where firebrands are likely to land or accumulate in a WUI environment. Further and more targeted analysis the differences in pyrolysis products for each material will likely yield further insight. 
    %\cite{MacLean1941}.

### 2022-05-25 thesis.tex
Updates to manuscripts
Updates to text an  manuscript naming
Updated results for manuscript4
Added CFD plot to manuscript 4
updated some formatting
completed conclusiosns

\submitdate{June 13, 2018}
\commencementyear{2019}

@@

\acknowledgements{
\textit{here you might thank the people that helped you, and possibly a funding source}
}

\dedication{(optional dedication)}

@@

% By including `''%!TEX root = thesis.tex` in the top of each included file,
% most LaTeX editors will automatically compile the main package when you are working
% within a file.

\include{introduction}
\include{litreview}
\include{manuscript1}
\include{manuscript2}
\include{manuscript3}
\include{results}
\include{conclusions}

>>>>

\submitdate{June 8, 2022}
\commencementyear{2022}

@@

\acknowledgements{
I have been fortunate to be perpetually surrounded by people that have given me an immense amount of support and guidance throughout my life and this has been especially true in my studies. This period of my life has been truly transformative and it would not have been possible without all of the wonderful people in my life. I owe the greatest thanks to my family whose unwavering support, encouragement, and guidance has been invaluable. I would not be who I am or where I am today if it was not for you all. It is impossible to put into words how much I appreciate everything you have done for me. To Heather, thank you for being my partner through the toughest parts of this journey, for asking so many great questions, and encouraging me to keep going so that I may reach my goals. I doubt I would have made it this far without you.

Thank you to Dr. David Blunck for giving me the opportunity to conduct this work. In the graduate school experience you have facilitated I have achieved things I did not know I was capable of. I have the deepest gratitude for your guidance, insight, and the example that you set. I owe a similar debt of gratitude to my fellow students....

This work was supported by the National Institute of Standards and Technology as part of project number 70NANB17H281.
}

%\dedication{(optional dedication)}

@@

% By including `''%!TEX root = thesis.tex` in the top of each included file,
% most LaTeX editors will automatically compile the main package when you are working
% within a file.
%\printnomenclature
\include{introduction}
\include{litreview}
\include{manuscript1}
\include{manuscript2}
\include{manuscript3}
\include{manuscript4}
%\include{results}
\include{conclusions}
